---
title: "Get Elements Of A GRM"
author: "Peter von Rohr"
date: "`r Sys.Date()`"
output: 
  html_document:
    mathjax: "http://cdn.mathjax.org/mathjax/latest/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Given the following problem. Suppose we have a large symmetric matrix of a certain dimension $n$ of which the lower triangular part is stored in a vector. We want to find all row- and column-indices of elements of the original matrix that fullfill a certain property. 

### Background
The genetic relationship matrix (GRM) between $n$ individuals is a symmetric matrix of dimension $n\times n$. Software programs that are used to compute the GRM such as GCTA store the lower triangular part of the matrix in a vector. In case we want to know pairs of individuals that have genetic relationship coefficients above a certain threshold, we need to know the row- and column-indices of the original matrix where the coefficients above a certain threshold occur in the original GRM. 

### Small example
Let us try to explain the problem based on a small example. Suppose the GRM between a small number of individuals has the following structure. For the purpose of this blog-post, we create a matrix based on random uniform numbers. Since the diagonals of a GRM are all larger than 1, we add 1 to the diagonal.


```{r GenerateGRM}
set.seed(4321)
nNrInd <- 6
mGrmP <- matrix(runif(n = nNrInd * nNrInd), ncol = nNrInd)
mGrm <- crossprod(mGrmP)/nNrInd
diag(mGrm) <- 1+diag(mGrm)
print(mGrm)
```

### First flattening the matrix into a vector
The lower triangular part of the above matrix can be stored in a vector as follows. 

```{r FlattenGrm}
vGrmLowTri <- vector(mode = "numeric", length = sum(1:nNrInd))
nVecIdx <- 1
for (nRowIdx in 1:nrow(mGrm)){
  for (nColIdx in 1:nRowIdx){
    vGrmLowTri[nVecIdx] <- mGrm[nRowIdx, nColIdx]
    nVecIdx <- nVecIdx + 1
  }
}
print(vGrmLowTri)
```

From the above, we can see that the order of the elements is to be taken row-wise, that means, in the outer loop we step through each row of the matrix and inside of each row, the inner loop runs from the first element up until the diagonal element. 

### Alternative to loops
R has internal functions to extract lower and upper triangular elements of a matrix. Those functions are `lower.tri()` and `upper.tri()` as arguments they take the matrix and a boolean flag which indicates whether the diagonal elements should be included or not. 

Since we are interested in the lower triangular of our matrix, it would seam obvious to use the function `lower.tri()`. 

```{r LowerTriMatrix}
lower.tri(mGrm, diag = TRUE)
```

The result of applying the function `lower.tri()` to our matrix is a boolean matrix of the same dimension as the input matrix where all elements of the lower triangular including the diagnal elements are TRUE and all other elements are FALSE. 

Using the boolean matrix that comes out of function `lower.tri()` as index selector for our original GRM matrix, flattens the lower triangular elements into a vector but the order is taken column-wise and not row-wise. As a consequence the resulting flattened vector is not the same as the vector that we obtained from our two iterated loops above. Therefore, it seams that our idea of using R's internal function `lower.tri()` for flattening the GRM matrix into a vector does not work. 


```{r FlattenRowWise}
mGrm[lower.tri(mGrm, diag = TRUE)]
```

### Saving our idea
Our original numeric matrix is symmetric and hence the elements above the diagonal are the same as below the diagonal. Further more the order of the elements is the same when running through the matrix row-wise below the diagonal as when running through the matrix column-wise above the diagonal. Hence, instead of using `lower.tri()`, we have to use `upper.tri()`. As shown below, we get the same flattened vector as from the two iterated loops. 

```{r FlattenColumnWise}
vFlatMatElem <- mGrm[upper.tri(mGrm, diag = TRUE)]
print(vFlatMatElem)
identical(vFlatMatElem, vGrmLowTri)
```

## Finding a given element in the matrix
Let us assume, we are given the flattened vector of the lower triangular part of our GRM matrix. Furthermore, we want to select some special components in the vector and want to trace back where they occured in the original matrix. That means we want to know the row and the column indices of the special component in the original matrix.

### Diagonal elements
One group of special elements are the diagonal elements of the original matrix. From our process of flattening the lower triangular part of our original matrix into a vector, we can observe that on a given row i, there are exactly i elements from column 1 until the diagonal element of row i. Based on that observation, the number of elements that are stored in the flattened vector up to the diagonal element of row i corresponds to the sum of all natural numbers from 1 up and including i. 

$$ \sum_{j=i}^i j$$

Computing that number for all rows in the matrix we can store all indices of the flattened vector corresponding to diagonal elements as follows. 

```{r DiagonalElementsIndex}
vFlatDiagIdx <- cumsum(1:nrow(mGrm))
print(vFlatDiagIdx)
```

From that index vector, all diagonal elements can be extracted by 

```{r ExtractDiagonalElement}
vFlatMatElem[vFlatDiagIdx]
```

Now it is easy to also get all the off-diagonal elements by

```{r ExtractOffDiagonals}
vFlatMatElem[-c(vFlatDiagIdx)]
```

### A particular element
Let us assume, we are interested in element 12 of the flattened vector. 

```{r Element12}
vFlatMatElem[12]
```

Our goal is to find out on which row and which column this element occurs in the original matrix. Our index vector of the diagonal elements already tells us that element 10 is the last element of the fourth row and element 15 is the last element of the fifth row. Element 12 must then be somewhere on the fifth row. We now want to compute where exactly that element occurs in the matrix. First we start with the row index. The row index corresponds to the index of our index vector of the diagonal element, where for the first time the element in the index vector is larger than the index of our special element. The column index is computed as the difference between the index of our special element (12) and the index of the diagonal element (10) of the previous row (4)

```{r GetMatRowIndex}
nElemIdx <- 12
nElementRow <- which(vFlatDiagIdx > nElemIdx)[1]
nElementCol <- nElemIdx-vFlatDiagIdx[nElementRow-1]
cat(" * Row: ", nElementRow, "\n * Col: ", nElementCol, "\n")
```

Checking our result

```{r CheckResult}
identical(vFlatMatElem[12], mGrm[nElementRow,nElementCol])
```

## Wrapping all up into a function
In the above sections we gave an outline of an algorithm that allows us to trace back elements from a flattened vector back to the original matrix. We want to wrap that up into a function, that takes as input, the flattened vector and the index of a special element. The function computes the row and the column indices and returns the respective element of the matrix.

```{r FindGrmElement}
findSingleGrmElement <- function(pnElementIdx, pvFlatVec){
  ### # compute the dimensionality of the original matrx
  nMatDim <- nGetMatDim(pnLenFlatVec = length(pvFlatVec))
  ### # compute vector of indices of diagonal elements
  vDiagIdx <- cumsum(1:nMatDim)
  ### # check whether pnElementIdx is a diagonal element, if yes, then special case
  if (any(vDiagIdx == pnElementIdx)) {
    nDiagIdx <- which(vDiagIdx == pnElementIdx)
    lResultList <- list(nRowIndex=nDiagIdx, nColIndex=nDiagIdx, nCoefficient = pvFlatVec[pnElementIdx])
  } else {
    ### # off-diagonal
    nElementRow <- which(vDiagIdx > pnElementIdx)[1]
    nElementCol <- pnElementIdx - vDiagIdx[nElementRow-1]
    lResultList <- list(nRowIndex=nElementRow, nColIndex=nElementCol, nCoefficient = pvFlatVec[pnElementIdx])
  }
  return(lResultList)
}
```

The function `findSingleGrmElement()` searches for a single element in the original matrix. The search is done only based on the element index and based on the flattened vector. In order to find our special element, we need a number of helper functions. The first helper function determines the dimension of the original matrix based on the length of the flattened vector. This can be done using the fact that the flattened vector contains all elements of the lower triangular part of the original matrix including the diagonal elements. Assuming that the original matrix has dimension $n\times n$, the total number of elements of the original matrix is $n^2$. The total number of elements below the diagonal without the diagonal elements is $(n^2-n)/2$. Adding the diagonal elements back in leads to 

$$(n^2-n)/2 + n = n*(n+1)/2$$ 

This expression corresponds to the length ($l$) of our flattened vector. Given the length of the flattened vector, we can compute the number of rows and columns (n) of our original matrix, based on the solution formula for quadratic equations. 

$$n = (-1 + \sqrt{1 + 8*l})/2 $$

```{r GetMatDim}
nGetMatDim <- function(pnLenFlatVec) {
  # pnLenFlatVec must a be positive integer
  if (pnLenFlatVec < 0) 
    stop(paste("Length of flattened vector must be positive, found: ", pnLenFlatVec))
  if (as.integer(pnLenFlatVec) - pnLenFlatVec != 0)
    stop(paste("Length of flattened vector must be an integer, found: ", pnLenFlatVec))
  # compute the matrix dimension
  nMatDimResult <- (sqrt(1 + 8*pnLenFlatVec) - 1) / 2
  # result must be integer
  if ((as.integer(nMatDimResult) - nMatDimResult) != 0 )
    stop(paste("Resulting matrix dimension not an integer:", nMatDimResult ))
  return(nMatDimResult)
}
```

